Line of Sight Lensing Mass Reconstructions in the Millennium Simulation

### 2016-06-05 doc/thesis_swe.tex
Thesis update. #98

% Used for .bib file
\newcommand{\mnras}{MNRAS}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJ}

% Used for footnotes
\makeatletter
\newcommand\footnoteref[1]{\protected@xdef\@thefnmark{\ref{#1}}\@footnotemark}
\makeatother

\raggedbottom

@@

\title{SPENCER THESIS - WORK IN PROGRESS}
%\thanks{A footnote to the article title}%

@@

%%==========================================================================================================================================
\begin{abstract}
\noindent (TEMPORARY) Recent evidence suggests that dark matter dominates over the normal matter we experience in everyday life, yet we are unable to directly map it as it does not absorb or emit light. Being able to predict high-resolution dark matter mass maps would be interesting in of itself, but from this we can also study structure formation and constrain the parameters of the universe. In this work, I develop the Pangloss model which approximates complex mass distributions with a simple dark matter halo model and then analyze how accurate its predictions can be using weak gravitational lensing data in the Millennium Simulation. I then do some exploratory work in estimating how feasible a large inference would be using both likelihood estimation and approximate Bayesian computation (ABC). I found that ...

% OLD: As dark matter does not absorb or emit light, its distribution in the universe must be inferred through indirect effects such as the gravitational lensing of distant galaxies. While most sources are only weakly lensed, the systematic alignment of background galaxies around a foreground lens can constrain the mass of the lens which is largely in the form of dark matter. In this paper, I have implemented a framework to reconstruct all of the mass along lines of sight using a  best-case dark matter halo model in which the halo mass is known. This framework is then used to make predictions of the weak lensing of 3,240 generated source galaxies through a 324 arcmin$^2$ field of the Millennium Simulation. The lensed source ellipticities are characterized by the ellipticity-ellipticity and galaxy-mass correlation functions and compared to the same statistic for the intrinsic and ray-traced ellipticities. In the ellipticity-ellipticity correlation function, I find that the framework systematically under predicts the shear power by an average factor of 2.2 and fails to capture correlation from dark matter structure at scales larger than 1 arcminute. The model predicted galaxy-mass correlation function is in agreement with the ray-traced statistic from scales 0.2 to 0.7 arcminutes, but systematically underpredicts shear power at scales larger than 0.7 arcminutes by an average factor of 1.2. Optimization of the framework code has reduced the mean CPU time per lensing prediction by 70\% to $24\pm5$ ms. Physical and computational shortcomings of the framework are discussed, as well as potential improvements for upcoming work.
\end{abstract}

%\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
                             % Classification Scheme.
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle

@@

In a universe teeming with galaxies and light, it came as a shock when 20\textsuperscript{th} century astronomers discovered that most of the mass in the universe is in fact dark; the `normal' matter made of atoms that we interact with in everyday life, called baryonic matter, accounts for less than 20\% of the mass in the observable universe \cite{planck}. The remaining mass takes the form of an exotic dark matter that does not absorb or emit light rendering it invisible to our telescopes. While this claim sounds bizarre, there has been an abundance of indirect evidence in recent decades for the existence of dark matter including the flattening of galaxy rotation curves \cite{rotation_curve}, velocity dispersions of galaxies \cite{zwicky}, acoustic peaks in the cosmic microwave background \cite{planck_2015}, and galaxy cluster collisions \cite{bullet_cluster}.

One of the most successful probes of dark matter has been gravitational lensing. The path of light from distant `background' galaxies is bent when traveling through regions of space containing large amounts of `foreground' mass. Light from different origins in a source galaxy is subject to different bending which results in a distortion of the galaxy image. As the foreground mass is known to be largely dark matter, gravitational lensing supplies a direct constraint on the mass of dark matter in that region \cite{mass_constraint}.

@@

Predicting accurate dark matter mass maps from weak lensing requires understanding the correlation between galaxies and dark matter structure. Numerous N-body simulations have predicted dark matter structure to have evolved from overdense regions in the early universe into `halos' of virialized matter that eventually collapsed from self-gravity (cite), where galaxies could then form in the large gravitational potentials of the halos. However, there is still extensive reserach being done on the correct density profile of the halos (isothermal spheres \cite{isothermal_spheres}, NFW \cite{nfw}, BMO \cite{bmo}, Einasto \cite{einasto}), identifying the stellar mass - halo mass relation \cite{smhr}, abundance matching \cite{reddick}, halo substructure \cite{substructure}, and the lensing by line-of-sight halos \cite{collet} and by dark matter structures not associated with galaxies \cite{mccully}. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of the weak lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation in order to predict the weakly lensed ellipticities of generated background sources. I then compare my predicted lensed ellipticities to the `true' lensing calculated by Hilbert et al. in \cite{ray_tracing} by ray-tracing light through the Millennium Simulation. By doing so, I attempt to answer the following questions:

\begin{enumerate}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model?
%\item What fraction of dark matter mass is in halos?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{enumerate}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Millennium Siulation structure
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/ms_structure.jpg}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION}
    \label{fig:ms_structure}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% FIGURE: Millennium Simulation Distribution plots (cosmic web, galaxy clusters with/without dark matter)
\begin{figure*}
    \centering
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_galaxies.jpg}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{ms_galaxies}
    \end{subfigure}
    ~
    %% Einstein Ring Figure
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_halos.jpg}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{ms_halos}
    \end{subfigure}
    \caption{}
    \label{ms_compare}
\end{figure*}

@@

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments pocketed with under-dense regions called voids, as shown in Figure \ref{fig:ms_structure} \cite{??}. While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter \cite{??}. This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures \cite{??}. A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure \ref{ms_compare}.

%However, these simulations also indicate that galaxies tend to form in the large gravitational potentials of dark matter over-dense regions of dark matter. While the exact relation between the distribution of galaxies and dark matter is not known, this m

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of ?? \cite{??}. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Filaments, Voids, and the Smooth-Component Correction} \label{scc}

@@

% With this, the proper volume of olume of the lightcone a given grid slice at redshift $z$ in a foreground catalog centered at ($\theta,\phi$) by integrating the above volume element over the solid angle $d\Omega$ and $\Delta z$ range of the redshift slice:
% \begin{align}
% V(z)&=\int_{V_z}dV_p\\ \nonumber
% &=d_H\int_\Omega\int_{z-\frac{\Delta z}{2}}^{z+\frac{\Delta z}{2}} \frac{d_M(z^\prime)^2}{E(z^\prime)(1+z^\prime)^3}dz^\prime d\Omega.
% \end{align}

@@

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics. The deviation in path is usually negligable, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when travelling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. A simplified cartoon of this process is shown in Figure ??.

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Gravitational Lensing Diagram
% \begin{figure*}
%     \centering
%     \begin{subfigure}[H]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/thesis/lens_cartoon.png}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{CAPTION}
%         \label{lens_cartoon}
%     \end{subfigure}
%     ~
%     %% Einstein Ring Figure
%     \begin{subfigure}[H]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/thesis/lens_diagram.png}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{CAPTION}
%         \label{lens_diagram}
%     \end{subfigure}
%     \caption{}
% \end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

\noindent where $\sum(\vec{r},z_l)$ is the projected surface mass density of the lensing mass and $\sum_{cr}(z_l,z_s)$ is the critical surface density for a given source redshift $z_s$ and lens redshift $z_l$ \cite{??}. The shearing of the source is most often described by the complex shear $\gamma$ defined to be
\begin{equation}\label{complex_shear}
\gamma=\gamma_1+i\gamma_2=|\gamma|\,e^{2i\varphi}
\end{equation}

\noindent where $|\gamma|$ is the magnitude of the shear and $\varphi$ is the orientation of the shear. While the intrinsic ellipticities of source galaxies are randomly oriented near the foreground mass before lensing, they will be systematically more aligned with the shear field after lensing.% Figure (??) demonstrates this process visually.

@@

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in \ref{bmo}. Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{kappa}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy \cite{??}. However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

@@

If all galaxies were spherical then the measurement of their ellipticities would give an exact description of the lensing done by foreground mass structures. Unfortunately, the intrinsic ellipticities of background galaxies contribute significant noise as we do not know how any individual galaxy ellipticity has been altered by lensing. However as two sources lensed by the same foreground mass will have their ellipticities distorted in a similar way, their lensed ellipticities should be correlated regardless of intrinsic orientation. The average correlation of a sufficient number of galaxy pairs should overcome the noise and lead to a detectable signal. Importantly, this implies that \textit{the correlation of galaxy ellipticities is a probe of the underlying dark matter mass distribution} \cite{??}.

@@

A similar statistical measure often used in weak lensing measurements is the galaxy-mass correlation function which, in this setting, measures the cross-correlation of galaxy ellipticities and halo locations. Simply put, the function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance and only has one component relevant for this work (see \cite{ng_corr} for details).

@@

To infer foreground dark matter mass structures using weak gravitational lensing, first a model of the relationship between foreground galaxies and dark matter must be established and robustly tested to see if, statistically, it makes the same lensing predictions of background sources as the true underlying dark matter structure. To do this, I built upon the publicly available \texttt{Pangloss}\footnote{\label{note1}https://github.com/drphilmarshall/Pangloss} framework used in Collett et al. (\cite{collett_marshall}) to reconstruct all the
mass along the lines of sight of generated background galaxies in the MS using a halo model with smooth-component correction. The lensing contribution of each halo is calculated, and the total lensing of the background galaxy is the sum of each halo contribution. The implementation of this process is detailed in the following sections.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{The Millennium Simulation} \label{ms}
\texttt{Pangloss} cannot be used to make dark matter mass maps using existing galaxy catalogs until it is tested on a simulated universe with known dark matter structure to determine how accurately and precisely it predicts the lensing of background sources. For this purpose, \texttt{Pangloss} was tested on galaxy catalogs from the Millennium Simulation, an N-body simulation consisting of over 10 billion dark matter `particles' each representing a billion solar masses and populated with about 20 million galaxies \cite{millennium_simulation}. The simulation contains baryonic and dark matter structure believed to be similar to our own universe, and uses cosmological parameters from WMAP 1\textsuperscript{st}-year data analysis which assumed a matter density of $\Omega_m=0.25$, a dark energy density of $\Omega_\Lambda=0.75$, and a Hubble constant of $H_0=73$ km$\,$s$^{-1}\,$Mpc$^{-1}$ (Cite). From the work of Hilbert et al. \cite{ray_tracing}, high resolution maps of the ray-traced shear and convergence of galaxy sources at $z=1.3857$ are publicly available. From these maps, an estimate of the `true' lensing of background galaxy ellipticities when traveling through the foreground mass of the Millennium Simulation can be calculated using Equation \eqref{lensing}.

@@

The volume and projected surface mass densities of each redshift slice are shown in Figures \ref{fig:volume_densities} and \ref{fig:surface_densities} respectively. In each plot, the halo density at each redshift slice is plotted in green, the smooth-component density in blue, and universe mean density in black, and a histogram of the galaxies per bin is shown in the background. In most redshift bins, the halo density is one to two orders of magnitude lower than than the universe mean density leading to a smooth-component density very close to that of the mean density. This seems to be in opposition to Pache et al. (year) \cite{pache} who found that halos constituted ??\% of simulation mass in ??. However, all mass densities are far below the critial density, plotted in red, and so well within the weak-lensing regime. The convergence distribution of each halo and 100 smooth-components is plotted in Figure \ref{fig:kappa_dist}.
%%------------------------------------------------------------------------------------------------------------------------------------------
%% Surface and Volume densities for smooth-component correction redshift slices
\begin{figure*}
    \centering
    % Volume
    \begin{subfigure}[H]{0.49\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/vol_densities.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{fig:volume_densities}
    \end{subfigure}
    ~
    % Surface
    \begin{subfigure}[H]{0.49\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/surf_densities.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{fig:surface_densities}
    \end{subfigure}
    \caption{}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

Instead of comparing the ray-traced and \texttt{Pangloss} predicted ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. As described in section \ref{corr_functions}, the ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-mass correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. Both correlation functions are used in this work to estimate how well the \texttt{Pangloss} framework models weak lensing by dark matter structures and are computed using the publicly available \texttt{TreeCorr} module written by Mike Jarvis\footnote{https://github.com/rmjarvis/TreeCorr}. Note that the ellipticity-ellipticity correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition and the more common Schneider definition \cite{schneider}, see Appendix \ref{corr_defs}.

@@

The result of a particular 9 arcmin$^2$ field is shown in Figure \ref{fig:kappamaps}, where the ray-traced convergence is shown on the left, the halo model convergence is in the center, and the ray-traced convergence subtracted from the halo model convergence is shown on the right. While particular foreground masses are almost always predicted, the currently implemented halo model over-predicts the convergence throughout most of the field; b. The halo model does not appear to accurately account for the voids between foreground masses and over-predicts the mass of most ... This could partially be a result of incorrect halo profiles, but it is apparent that the smooth-component correction is not adequate to ...

>>>>

% Used for .bib file
\newcommand{\mnras}{MNRAS}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJ}
\newcommand{\jcap}{JCAP}
\newcommand{\aj}{AJ}
\newcommand{\apjs}{ApJS}

\makeatletter
\newcommand\footnoteref[1]{\protected@xdef\@thefnmark{\ref{#1}}\@footnotemark} % Used for footnotes
\def\p@subsection{} % Fixes section references
\makeatother

% \raggedbottom

@@

\title{Line of Sight Lensing Mass Reconstructions in the Millennium Simulation}
%\thanks{A footnote to the article title}%

@@

%%==========================================================================================================================================
\begin{abstract}
\noindent 
% (TEMPORARY) Recent evidence suggests that dark matter dominates over the normal matter we experience in everyday life, yet we are unable to directly map it as it does not absorb or emit light. Being able to predict high-resolution dark matter mass maps would be interesting in of itself, but from this we can also study structure formation and constrain the parameters of the universe. In this work, I develop the Pangloss model which approximates complex mass distributions with a simple dark matter halo model and then analyze how accurate its predictions can be using weak gravitational lensing data in the Millennium Simulation. I then do some exploratory work in estimating how feasible a large inference would be using both likelihood estimation and approximate Bayesian computation (ABC). I found that ...
As dark matter does not absorb or emit light, its distribution in the universe must be inferred through indirect effects such as the gravitational lensing of distant galaxies. While most sources are only weakly lensed, the systematic alignment of background galaxies around a foreground lens can constrain the mass of the lens which is largely dark matter. In this thesis, I have implemented a framework to reconstruct lensing mass along lines of sight in a 144 arcmin$^2$ field of the Millennium Simulation by predicting the weak lensing of 1440 generated source galaxies using a best-case dark matter halo model in which the halo mass is known. The lensed source ellipticities are characterized by the ellipticity-ellipticity and galaxy-mass correlation functions and compared to the same statistic using the `true' ray-traced ellipticities. In the ellipticity-ellipticity correlation function, I find that the halo model systematically underpredicts the correlation on scales above 0.2 arcminutes by a mean NRMSE of 0.22, although this can be decreased by including more halo contributions. The model predicted galaxy-mass correlation function is in agreement with the ray-traced statistic on scales from 0.2 to 2 arcminutes but overpredicts the correlation on scales below 0.2 arcminutes due to strong lensing effects with a mean NRMSE of 0.36. Both best-case statistics were found to be well approximated by using only the most `relevant' halos, which on average only needed the 300 and 70 most relevant halos to predict the correlation functions with a NRMSE below 5\% and 10\% respectively. The feasibility of using the halo model to infer hyperparameters of the Millennium Simulation with either maximum likelihood estimation or approximate Bayesian computation is explored, with preliminary results favoring likelihood estimation. Optimization of the framework code has reduced the mean CPU time per lensing prediction by 59\% to $57\pm8$ ms, although this is approximately two orders of magnitude slower than required for a planned scale-up to a 100 arcmin$^2$ field. Physical and computational limitations of the framework are discussed, as well as potential improvements for future work.
\end{abstract}

\maketitle

@@

In a universe teeming with galaxies and light, it came as a shock when 20\textsuperscript{th} century astronomers discovered that most of the mass in the universe is in fact dark; the `normal' matter made of atoms that we interact with in everyday life, called baryonic matter, accounts for less than 20\% of the mass in the observable universe \cite{planck_2015}. The remaining mass takes the form of an exotic dark matter that does not absorb or emit light rendering it invisible to our telescopes. While this claim sounds bizarre, there has been an abundance of indirect evidence in recent decades for the existence of dark matter including the flattening of galaxy rotation curves \cite{rotation_curve}, velocity dispersions of galaxies \cite{zwicky}, the power spectrum of the cosmic microwave background radiation \cite{planck_2015}, acoustic peaks in the correlation function of luminous red galaxies \cite{bao_detection}, and galaxy cluster collisions \cite{bullet_cluster}.

One of the most successful probes of dark matter has been gravitational lensing. The path of light from distant `background' galaxies is bent when traveling through regions of space containing large amounts of `foreground' mass. Light from different origins in a source galaxy is subject to different bending which results in a distortion of the galaxy image. As the foreground mass is known to be largely dark matter, gravitational lensing supplies a direct constraint on the mass of dark matter in that region \cite{dark_matter_probe}.

@@

Predicting accurate dark matter mass maps from weak lensing requires understanding the correlation between galaxies and dark matter structure. Numerous N-body simulations have predicted dark matter structure to have evolved from overdense regions in the early universe into `halos' of virialized matter that eventually collapsed from self-gravity \cite{halo_formation}, where galaxies could then form in the large gravitational potentials of the halos. However, there is still extensive reserach being done on the correct density profile of the halos (isothermal spheres \cite{isothermal_spheres}, NFW \cite{nfw}, BMO \cite{nfw_bmo}, Einasto \cite{einasto}), identifying the stellar mass - halo mass relation \cite{smhr}, abundance matching \cite{abundance_matching}, halo substructure \cite{halo_substructure}, and the lensing by line-of-sight halos \cite{collett} and by dark matter structures not associated with galaxies \cite{mccully}. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of the weak lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

Guided by these results, in this thesis I build upon the work of Collett et al. in \cite{collett} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation in order to predict the weakly lensed ellipticities of generated background sources. I then compare my predicted lensed ellipticities to the `true' lensing calculated by Hilbert et al. in \cite{ray_tracing} by ray-tracing light through the Millennium Simulation. By doing so, I attempt to answer the following questions:

\begin{itemize}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{itemize}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Millennium Siulation structure
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/ms_structure.jpg}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{\cite{ms_figures}}
    \label{fig:ms_structure}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% FIGURE: Millennium Simulation Distribution plots (cosmic web, galaxy clusters with/without dark matter)
\begin{figure*}
    \centering
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_galaxies.jpg}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{ms_galaxies}
    \end{subfigure}
    ~
    %% Einstein Ring Figure
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_halos.jpg}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{ms_halos}
    \end{subfigure}
    \caption{\cite{ms_figures}}
    \label{ms_compare}
\end{figure*}

@@

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments pocketed with under-dense regions called voids, as shown in Figure \ref{fig:ms_structure} \cite{ms_figures}. While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter \cite{halo_formation}. This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures \cite{abundance_matching}. A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure \ref{ms_compare}.

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of between 300,000 and 400,000 light-yeras \cite{milky_way_halo}. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Filaments, Voids, and the Smooth-Component Correction}\label{scc}

@@

@@

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics. The deviation in path is usually negligable, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when travelling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. %A simplified cartoon of this process is shown in Figure \ref{fig:lens_diagram}.

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Gravitational Lensing Diagram
% \begin{figure}
%     \centering
%     \includegraphics[width=0.475\textwidth]{figs-swe/thesis/lensing_diagram.jpg}
%     \captionsetup{justification=raggedright,singlelinecheck=false}
%     \caption{CAPTION, Illustration by Ian Worpole (www.ianworpole.com)}
%     \label{fig:lens_diagram}
% \end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

\noindent where $\sum(\vec{r},z_l)$ is the projected surface mass density of the lensing mass and $\sum_{cr}(z_l,z_s)$ is the critical surface density for a given source redshift $z_s$ and lens redshift $z_l$ \cite{schneider}. The shearing of the source is most often described by the complex shear $\gamma$ defined to be
\begin{equation}\label{complex_shear}
\gamma=\gamma_1+i\gamma_2=|\gamma|\,e^{2i\varphi}
\end{equation}

\noindent where $|\gamma|$ is the magnitude of the shear and $\varphi$ is the orientation of the shear. While the intrinsic ellipticities of source galaxies are randomly oriented near the foreground mass before lensing, they will be systematically more aligned with the shear field after lensing.

@@

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in \ref{bmo}. Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{convergence}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy \cite{mass_degeneracy}. However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

@@

If all galaxies were spherical then the measurement of their ellipticities would give an exact description of the lensing done by foreground mass structures. Unfortunately, the intrinsic ellipticities of background galaxies contribute significant noise as we do not know how any individual galaxy ellipticity has been altered by lensing. However as two sources lensed by the same foreground mass will have their ellipticities distorted in a similar way, their lensed ellipticities should be correlated regardless of intrinsic orientation. The average correlation of a sufficient number of galaxy pairs should overcome the noise and lead to a detectable signal. Importantly, this implies that \textit{the correlation of galaxy ellipticities is a probe of the underlying dark matter mass distribution} \cite{dark_matter_probe}.

@@

A similar statistical measure often used in weak lensing measurements is the galaxy-mass correlation function which, in this setting, measures the cross-correlation of galaxy ellipticities and halo locations. Simply put, the function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance and only has one component relevant for this work (see \cite{gm_corr} for details).

@@

To infer foreground dark matter mass structures using weak gravitational lensing, first a model of the relationship between foreground galaxies and dark matter must be established and robustly tested to see if, statistically, it makes the same lensing predictions of background sources as the true underlying dark matter structure. To do this, I built upon the publicly available \texttt{Pangloss}\footnote{\label{note1}https://github.com/drphilmarshall/Pangloss} framework used in Collett et al. (\cite{collett}) to reconstruct all the
mass along the lines of sight of generated background galaxies in the MS using a halo model with smooth-component correction. The lensing contribution of each halo is calculated, and the total lensing of the background galaxy is the sum of each halo contribution. The implementation of this process is detailed in the following sections.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{The Millennium Simulation} \label{ms}
\texttt{Pangloss} cannot be used to make dark matter mass maps using existing galaxy catalogs until it is tested on a simulated universe with known dark matter structure to determine how accurately and precisely it predicts the lensing of background sources. For this purpose, \texttt{Pangloss} was tested on galaxy catalogs from the Millennium Simulation, an N-body simulation consisting of over 10 billion dark matter `particles' each representing a billion solar masses and populated with about 20 million galaxies \cite{millennium_simulation}. The simulation contains baryonic and dark matter structure believed to be similar to our own universe, and uses cosmological parameters from WMAP 1\textsuperscript{st}-year data analysis which assumed a matter density of $\Omega_m=0.25$, a dark energy density of $\Omega_\Lambda=0.75$, and a Hubble constant of $H_0=73$ km$\,$s$^{-1}\,$Mpc$^{-1}$ \cite{wmap1}. From the work of Hilbert et al. \cite{ray_tracing}, high resolution maps of the ray-traced shear and convergence of galaxy sources at $z=1.3857$ are publicly available. From these maps, an estimate of the `true' lensing of background galaxy ellipticities when traveling through the foreground mass of the Millennium Simulation can be calculated using Equation \eqref{lensing}.

@@

The volume and projected surface mass densities of each redshift slice are shown in Figures \ref{fig:volume_densities} and \ref{fig:surface_densities} respectively. In each plot, the halo density at each redshift slice is plotted in green, the smooth-component density in blue, and universe mean density in black, and a histogram of the galaxies per bin is shown in the background. In most redshift bins, the halo density is one to two orders of magnitude lower than than the universe mean density leading to a smooth-component density very close to that of the mean density. However, all mass densities are far below the critial density, plotted in red, and so well within the weak-lensing regime. The convergence distribution of each halo and 100 smooth-components is plotted in Figure \ref{fig:kappa_dist}.
% This seems to be in opposition to Pace et al. in \cite{pace}, which found that halos constituted ??\% of simulation mass in ??.
%%------------------------------------------------------------------------------------------------------------------------------------------
%% Surface and Volume densities for smooth-component correction redshift slices
\begin{figure*}
    \centering
    % Volume
    \begin{subfigure}[H]{0.49\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/vol_densities.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{fig:volume_densities}
    \end{subfigure}
    ~
    % Surface
    \begin{subfigure}[H]{0.49\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/surf_densities.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{fig:surface_densities}
    \end{subfigure}
    \caption{}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

Instead of comparing the ray-traced and \texttt{Pangloss} predicted ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. As described in section \ref{corr_functions}, the ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-mass correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. Both correlation functions are used in this work to estimate how well the \texttt{Pangloss} framework models weak lensing by dark matter structures and are computed using the publicly available \texttt{TreeCorr} module written by Mike Jarvis\footnote{https://github.com/rmjarvis/TreeCorr}. Note that the ellipticity-ellipticity correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition \cite{jarvis} and the more common Schneider definition \cite{schneider}, see Appendix \ref{corr_defs}.

@@

The result of a particular 9 arcmin$^2$ field is shown in Figure \ref{fig:kappamaps}, where the ray-traced convergence is shown on the left, the halo model convergence is in the center, and the ray-traced convergence subtracted from the halo model convergence is shown on the right. While particular foreground objects are almost always predicted, the currently implemented halo model over-predicts the convergence throughout most of the field; (more). The halo model does not appear to accurately account for the voids between foreground masses and over-predicts the mass of most ... This could partially be a result of incorrect halo profiles, but it is apparent that the smooth-component correction is not adequate to ... (rework)
